---
title: "Crash Analysis Around Crosswalk Removal"
author: "Benedict Leonardi"
execute:
  eval: false
  echo: true
---

## Overview

After nearly getting hit a few times walking the area around historic Newport, KY, I was indignant enough to call local officials and try to understand what was going on.

This analysis evaluates changes in crash frequency following the removal of a crosswalk due to an accident which took out a pole which lit the intersection. The city of Newport, KY wisely and totally unsurpisingly decided that if a car lost control because there a road was overbuilt, the solution is to remove anything that would slow drivers in the future. The data use is pulled from [KYTC public data](http://crashinformationky.org/AdvancedSearch), filtered to (first) the City of Newport, then to a small bounding box around the affected intersection.

We use a Poisson regression to model monthly crash counts as a function of time, citywide crash activity, and whether the crash occurred *after* crosswalk removal.

## Setup

```{r}
library(tidyverse)
library(sf)
library(lubridate)

# Load and clean data
all_crashes_df <- read_csv("all_crashes.csv") %>%
  mutate(collision_date = mdy_hms(collisiondate)) %>%
  mutate(
    collision_hr = str_extract(collisiontime, "[0-9]{2}"),
    collision_min = str_extract(collisiontime, "[0-9]{2}$")
    # collision_date = date(collisiondate)
  )

# Convert to spatial data
all_crashes_sf <- all_crashes_df %>%
  filter(!is.na(latitude) & !is.na(longitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# Define area of interest (intersection)
bbox <- st_bbox(c(xmin = -84.493340, ymin = 39.096533,
                  xmax = -84.490958, ymax = 39.098398), crs = 4326)
```

## Visualization

```{r}
tmp_sf <- all_crashes_sf %>%
    st_filter(st_as_sfc(bbox)) %>%
    mutate(after_remove_xwalk = collision_date > ymd("2022-11-01"))
mapview::mapview(tmp_sf, zcol = "after_remove_xwalk")
```

## Modeling

```{r}
run_model <- function(date_threshold) {
  all_crashes_sf_filtered <- all_crashes_sf %>%
    st_filter(st_as_sfc(bbox)) %>%
    mutate(after_remove_xwalk = collision_date > ymd(date_threshold))
  
  crash_counts <- all_crashes_sf_filtered %>%
    st_drop_geometry() %>%
    mutate(month = floor_date(collision_date, "month")) %>%
    group_by(month, after_remove_xwalk) %>%
    summarise(crash_count = n(), .groups = "drop") %>%
    arrange(month) %>%
    mutate(month_index = as.numeric(difftime(month, min(month), units = "weeks")) / 4.345)
  
  citywide_counts <- all_crashes_sf %>%
    st_drop_geometry() %>%
    mutate(month = floor_date(collision_date, "month")) %>%
    group_by(month) %>%
    summarise(city_crash_count = n(), .groups = "drop")
  
  crash_counts <- crash_counts %>%
    left_join(citywide_counts, by = "month")
  
  glm(crash_count ~ after_remove_xwalk + month_index + log(city_crash_count),
      family = poisson(), data = crash_counts) %>% summary()
}

run_model("2022-11-01")
run_model("2024-02-01")
```

## Results

Because Google Maps streetview information was hard to come by and I could not pinpoint which accident in particular took out the pole necesarry to support the crosswalk overhead signal, I used a few dates:

### Crosswalk Removed: Nov 2022

```         
Coefficient: after_remove_xwalk = 0.31

Interpretation: exp(0.31) ≈ 1.36

Conclusion: ~36% increase in crashes after removal.
```

### Crosswalk Removed: Feb 2024

```         
Coefficient: after_remove_xwalk = 0.50

Interpretation: exp(0.50) ≈ 1.64

Conclusion: ~64% increase in crashes after removal.
```

### Notes

```         
All results are statistically significant (p < 0.001).

The magnitude of estimated increase varies depending on the assumed removal date.

Monthly time trend and citywide crash volume are controlled in the model.
```
